---
title: "Working with tameDP"
author: "A.Chafetz"
date: "4/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## tameDP package overview

Over the last few years, the SI team has maintained a package for working with the Data Pack called [`tameDP`](https://github.com/USAID-OHA-SI/tameDP). The principle idea is to get a tidy dataset out of the PSNUxIM tab to work with it. There are a numbe of useful features including adding in the additional HTS modalities (from VMMC, PMTCT, TB, and index), breaking out the `indicatorcode` column into discrete usable/filterable pieces, providing mechanism/parnter info, and creating a HTS_TST_POS indicator. 

```{r}
#install
install.packages("remotes")
remotes::install_github("USAID-OHA-SI/tameDP")

#load library
library(tameDP)
```

## Extracting the data
The main function is pretty easy to run, providing just the file path to the data pack you're working with.

```{r}
filepath <- "tools/COP21_DataPack_Saturn.xlsx"

df_sat <- tame_dp(filepath)
```
The most time consuming piece is hitting DATIM to connect the `mech_code` to other mechanism information. To speed things up, you can stop the mechanism mapping, by setting the `map_names` parameter to `FALSE`. 

```{r}
df_sat <- tame_dp(filepath, map_names = FALSE)

```
This is definitely advisable if you are ingesting multiple data packs so you don't hit DATIM umpteen times.

```{r}
#load package
  library(purrr)

#identify all the Data Pack files
  files <- list.files("tools", full.names = TRUE)

#read in all DPs and combine into one data frame
  df_all <- map_dfr(.x = files,
                    .f = ~ tame_dp(.x, map_names = FALSE))
                    
#apply mech_name and primepartner name, funding agency and OU from DATIM
  df_all <- get_names(df_all)
```

## Example query easy to deal with in tidy format
To see how useful it can be using `tameDP` over Excel, let's try to answer the following question: what districts (PSNUs) have the highest share newly identified positives from the community?

```{r}
library(dplyr)

df_sat %>% 
  filter(indicator == "HTS_TST_POS",
                disagg != "KeyPop",
                mech_code != "00000") %>% 
  mutate(source = ifelse(stringr::str_detect(modality, "Mod"), "Comm", "Fac")) %>%
  count(psnu, source, wt = targets, name = "targets") %>% 
  group_by(psnu) %>% 
  mutate(share = targets/sum(targets, na.rm = TRUE)) %>% 
  ungroup() %>% 
  filter(source == "Comm") %>% 
  arrange(desc(share)) %>% 
  print(n = Inf)
```


## SGAC alternative package

The SGAC systems team, who has taken ownership of the Data Pack the last few years, has a [package they use for Data Pack](https://github.com/pepfar-datim/datapackr) development which can also be used for getting a somewhat tidy output back. An additioonal benefit is you are also able to extract data from the tabs leading up to the PSNUxIM tab.

```{r}
#install
install.packages("remotes")
remotes::install_github("pepfar-datim/datapackr")

#load library
library(datapackr)

#provide DATIM credentials
loginToDATIM()

#"unpack" the data pack, creating a nested list of objects
d <- unPackTool(filepath)

#extract the PSNUxIM tab
dp_tidy <-  d$data$SNUxIM

```