---
title: Data Transformation
date: 2020-05-11
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This exercise is devided into 2 parts and designed to work you through data transformation processes.

1. Subsetting your data

2. Transform your data


## Load required libraries

```{r}
#Dependencies
  library(tidyverse)
  library(ICPIutilities)
```


## import row data
```{r}
  dataset_url <- "https://media.githubusercontent.com/media/ICPI/TrainingDataset/master/Output/MER_Structured_TRAINING_Datasets_PSNU_IM_FY18-20_20200214_v1_1.txt"
  
  df <- read_msd(dataset_url, save_rds = FALSE)
  
  
  df_dplyr <- df %>% 
    filter(indicator %in% c("TX_CURR", "TX_NEW", "HTS_TST", "HTS_TST_POS"),
           operatingunit %in% c("Saturn", "Jupiter"),
           standardizeddisaggregate == "Total Numerator") %>% 
    select(operatingunit, psnu, fundingagency, mech_name, indicator, fiscal_year:cumulative)
  
  set.seed(42)
  
  extra_zeros <- df_dplyr %>% 
    sample_n(10) %>% 
    mutate_if(is.double, ~ 0)
  
  df_dplyr <- bind_rows(df_dplyr, extra_zeros)
  
```

## Subsetting your data

Subset the original dataset by filtering down indicators and operatingunits.

You only need to keep 4 indicators: TX_CURR, TX_NEW, HTS_TST, HTS_TST_POS 

and 2 operating units: Saturn, Jupiter


```{r}
df <- df %>%
  filter(
    indicator %in% c("TX_CURR", "TX_NEW", "HTS_TST", "HTS_TST_POS"),
    operatingunit %in% c("Saturn", "Jupiter")
  )
```


Subset your dataset again and only keep columns of interest, about 12 of them

You will need to select: operatingunit, psnu, fundingagency, mech_name, indicator, and all columns between fiscal_year and cumulative


```{r}
df <- df %>% 
  select(operatingunit, psnu, fundingagency, mech_name, indicator, fiscal_year:cumulative)
```


Try to rename some of the columns: operatingunit, fundingagency and cumulative


```{r}
df <- df %>% 
  rename(
    operating_unit = operatingunit,
    funding_agency = fundingagency,
    cumulative_result = cumulative
  )
```


## Transform you data

Re-calculate the cumulative results values and make sure results makes sense


```{r}
df <- df %>% 
  rowwise() %>% 
  mutate(cumulative_result = sum(qtr1, qtr2, qtr3, qtr4, na.rm = TRUE))
```

Make sure zero in the cumulative result field is NA

```{r}
df <- df %>%
  mutate(cumulative_result = ifelse(cumulative_result == 0, NA, cumulative_result))
```

Add a new column to hold the average result of all 4 quarters

Make sure to exlude NAs. Does this average result make sense to you?


```{r}
df <- df %>%
  mutate(
    average_result = rowMeans(cbind(qtr1, qtr2, qtr3, qtr4), na.rm = TRUE),
    average_result = as.integer(average_result))
```

Get a list of IM and Indicator with achievement rates greater than 100 %

Focus on one operating unit and the current fiscal year

```{r}
df %>% 
  mutate(achievement_rate = round(cumulative_result / targets * 100, 1)) %>%   filter(fiscal_year == 2020, operating_unit == 'Saturn', achievement_rate > 100) %>% 
  distinct(mech_name, indicator) %>% 
  pull(mech_name)

```