---
title: "transform_data"
author: "jdavis"
date: "5/11/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ICPIutilities)

dataset_url <- "https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-04-27/dplyr_exercise.csv"
df <- read_csv(dataset_url)

full_dataset_url <- "https://media.githubusercontent.com/media/ICPI/TrainingDataset/master/Output/MER_Structured_TRAINING_Datasets_PSNU_IM_FY18-20_20200214_v1_1.txt"
df_full <- read_msd(full_dataset_url, save_rds = FALSE)

#have a look
glimpse(df)


```
Running order for 5.11


1. Review `filter`
2. `Select`
3. `Mutate`




dplyr::select
`select` extracts columns by name. 

```{r}
glimpse(df_full)

df_full %>% select(operatingunit, psnu, fundingagency, fiscal_year, targets) %>% 
  glimpse()

## change the order of the vars within your select statement

df_full %>% select(targets, fiscal_year, operatingunit, psnu, fundingagency) %>% 
  glimpse()

```

##helpers part 1 : ranges and `ends_with`
Pretty straight forward right? If you have a lot of things you want to select, it can be a pain to list everything out in a select statement, in that case there are a couple of options. 
```{r}
## use a start_col:end_col, or a range
df_full %>% select(operatingunit:targets) %>% 
  glimpse()


##combine ranges

df_full %>%
  select(operatingunit:fundingagency, standardizeddisaggregate:targets) %>% 
  glimpse()

##you can also deselect columns by adding a - in front of them : note use of parans

df_full %>%
  select(-region, -regionuid, -(snuprioritization:disaggregate)) %>% 
  glimpse()

## you can even deselect a chunk and re-add a column
glimpse(df_full)

df_full %>%
  select(-(snuprioritization:modality), standardizeddisaggregate, indicator) %>% 
  glimpse()

## if you have data that has column names of a similar struture, you can use partial matching
df_full %>%
  select(operatingunit, ends_with("uid")) %>%
  glimpse()


## cool, but not very helpful, we really want to do the reverse
df_full %>%
  select(-ends_with("uid")) %>% 
  glimpse()

```


MUTATE
Let's create new columns!
`mutate` has an almost endless number of options, we'll try and cover a few here that we've found to be useful with PEPFAR data

One of the simplest options is a calculation based on values in other columns. 

```{r}
glimpse(df)

df %>%
  filter(fiscal_year == 2020) %>% 
  mutate(qtr_avg = cumulative/4) %>%
  glimpse() %>% 
  view()

## make it look a little nicer

df %>%
  filter(fiscal_year == 2020) %>% 
  mutate(achievement = cumulative/targets) %>%
  view()

df %>% 
  mutate(achievement = round(cumulative/targets,2)) %>%
  filter(fiscal_year == 2020) %>% 
  view()


## more useful still
df %>%
  filter(fiscal_year == 2019,
         mech_name == "Ksr",
         indicator %in% c("HTS_TST_POS")) %>%
  mutate(achievement = round(cumulative/targets,2)) %>%
  view()

```

`mutate` can be used to create a new column, or replace an exsisting column by keeping the same name
Below is a really good thing to add when working with the MSD

```{r}
df %>% distinct(fundingagency)
glimpse(df)

df %>% 
  mutate(fundingagency = str_remove(fundingagency, "HHS/")) %>% 
  count(fundingagency)

df %>% 
  mutate(fundingagency = ifelse())

df %>% 
  mutate(fundingagency = str_remove(fundingagency, "HHS/"),
           fundingagency = factor(fundingagency, c("USAID", "CDC", "DOD"))) %>% 
  count(fundingagency)

```

We're going to tiptoe into condtions now. There is a lot to cover so we'll start slow.
A very simple but powerful function to use with `mutate` is `if_else()`. With `if_else()`, you first specify a logical statement, afterwards what needs to happen if the statement returns TRUE, and lastly what needs to happen if it’s FALSE. ([What is the difference between ifelse() and if_else?](https://medium.com/@HollyEmblem/chaining-if-else-statements-using-the-superheroes-dataset-with-dplyr-b13121777b4))

```{r}

# let's make some groups, shall we?
distinct(df, fundingagency)

## usaid vs others

df %>% 
  mutate(fundingagency = if_else(fundingagency == "USAID", "USAID", "Other")) %>% 
  count(fundingagency)
## note it dropped dedups

## a little nicer

df %>% 
  mutate(fundingagency = if_else(fundingagency == "USAID", "USAID", "Other"),
         fundingagency = factor(fundingagency, c("USAID", "Other"))) %>% 
  count(fundingagency)


## let's try an example with a logical operator
## how many mechs had targets for TX_CURR for USAID in 2019?
df %>% distinct(mech_name, fundingagency, fiscal_year)

df %>% distinct(mech_name, fundingagency, fiscal_year)
## not helpful

# a bit better
df %>% 
  filter(fundingagency == "USAID",
         indicator == "TX_CURR",
         fiscal_year == 2019) %>%
  mutate(no_targets = if_else(targets>0, "yes", "no")) %>%
  distinct(mech_name)


#another way
df %>% 
  filter(fundingagency == "USAID",
         indicator == "TX_CURR",
         fiscal_year == 2019) %>%
  mutate(no_targets = ifelse(targets>0, "yes", "no")) %>%
  distinct(no_targets, mech_name)


# an example of `case_when`
# case_when is useful for when you have more than one category (use if_else, or ifelse for that)
# the structure below is mutate(new_var = case_when(old_var == "old_var_value" ~ "value_you_want_new_var_to_have"))
# Important to remember `TRUE ~ old_var` at the end; this tells it to take the value of the old var except for what
## you've specified above

df %>% 
  mutate(tech_areas = case_when(indicator == "HTS_TST" ~ "Testing",
                                indicator == "HTS_TST_POS" ~ "Testing",
                                indicator == "TX_NEW" ~ "Treatment",
                                indicator == "TX_CURR" ~ "Treatment",
                                TRUE ~ indicator)) %>% 
  distinct(indicator, tech_areas)





```








