---
title: "transform_data"
author: "jdavis"
date: "5/11/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

MUTATE
Let's create new columns!
`mutate` has an almost endless number of options, we'll try and cover a few here that we've found to be useful with PEPFAR data

One of the simplest options is a calculation based on values in other columns. In the sample code, we’re changing the sleep data from data measured in hours to minutes.

```{r}
glimpse(df)

##let's create a flag variable for non-missing fy20 targets


## what about something more useful


## make it look a little nicer


## more useful still


```

`mutate` can be used to create a new column, or replace an exsisting column by keeping the same name
Below is a really good thing to add when working with the MSD

```{r}



```

We're going to tiptoe into condtions now. There is a lot to cover so we'll start slow.
A very simple but powerful function to use with `mutate` is `if_else()`. With `if_else()`, you first specify a logical statement, afterwards what needs to happen if the statement returns TRUE, and lastly what needs to happen if it’s FALSE. ([What is the difference between ifelse() and if_else?](https://medium.com/@HollyEmblem/chaining-if-else-statements-using-the-superheroes-dataset-with-dplyr-b13121777b4))

```{r}

# let's make some groups, shall we?



## note it dropped dedups
## usaid vs others



## make it a little nicer


## let's try an example with a logical operator
## how many mechs had targets for TX_CURR for USAID in 2019?

df %>% distinct(mech_name, fundingagency, fiscal_year)
## not helpful



```




























dplyr::select
`select` extracts columns by name. 

```{r}
glimpse(df)

df %>% select(operatingunit, psnu, fundingagency, standardizeddisaggregate, fiscal_year, targets) %>% 
  glimpse()

## change the order of the vars within your select statement
df %>% select(targets, fiscal_year, operatingunit, psnu, fundingagency, standardizeddisaggregate) %>% 
  glimpse()

```

##helpers part 1 : ranges and `ends_with`
Pretty straight forward right? If you have a lot of things you want to select, it can be a pain to list everything out in a select statement, in that case there are a couple of options. 
```{r}
## use a start_col:end_col, or a range

df %>% select(operatingunit:targets) %>% 
  glimpse()

##combine ranges
df %>% select(operatingunit:fundingagency, standardizeddisaggregate:targets)

##you can also deselect columns by adding a - in front of them : note use of parans
df %>% select(-region, -regionuid, -(snuprioritization:disaggregate))

## you can even deselect a chunk and re-add a column
df %>% select(-(snuprioritization:modality), standardizeddisaggregate, indicator) %>% 
  glimpse()

## if you have data that has column names of a similar struture, you can use partial matching
df %>% select(operatingunit, ends_with("uid")) %>%
  glimpse()

## cool, but not very helpful, we really want to do the reverse
df %>% select(-ends_with("uid")) %>% 
  glimpse()

```

## select_if
```{r}
#to do

```

## rename and rename_all
Next, we're going to talk about how to rename columns. This is really easy with `dplyr` and useful when working with the MSD as a lot of times you'll want to rename the vars in your df for display/viz purposes. We're all introducing the `_all` modifier for the dplyr functions here. Note that these are not helpers, they are standalone functions, just that the role they serve is a lot like the helpers. `rename` is pretty straightforward: `rename(new_name = old_name)` (note, if you are like me you will get that backwards a seemingly unbelievable number of times)
```{r}

df <- df %>% 
  rename()



```


##helpers part 2 : contains, starts_with, ends_with
This is where the helpers start to get neat, `dplyr` if you have a lot of columns with similar structure you can use some of partial matching to choose what you want.
```{r}
df %>% select()

```










##running list of helpers for the facilitation team to help josh figure out where to include
```{r}
one_col:another_col (range)
-c("one_col", "another_col") (except)
starts_with()
ends_with()
contains()
matches()
one_of()
## boolean
<
>
==
<=
>=
!=
%in%
is.na()
!is.na()




```






















