---
title: "transform_data"
author: "jdavis"
date: "5/11/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ICPIutilities)

dataset_url <- "https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-04-27/dplyr_exercise.csv"

df <- read_csv(dataset_url)

#have a look
glimpse(df)


```
Running order for 5.11


1. Review `filter`
2. `Select`
3. `Mutate`




dplyr::select
`select` extracts columns by name. 

```{r}
glimpse(df)



## change the order of the vars within your select statement


```

##helpers part 1 : ranges and `ends_with`
Pretty straight forward right? If you have a lot of things you want to select, it can be a pain to list everything out in a select statement, in that case there are a couple of options. 
```{r}
## use a start_col:end_col, or a range



##combine ranges

##you can also deselect columns by adding a - in front of them : note use of parans

## you can even deselect a chunk and re-add a column


## if you have data that has column names of a similar struture, you can use partial matching


## cool, but not very helpful, we really want to do the reverse


```


MUTATE
Let's create new columns!
`mutate` has an almost endless number of options, we'll try and cover a few here that we've found to be useful with PEPFAR data

One of the simplest options is a calculation based on values in other columns. 

```{r}
glimpse(df)

##let's create a flag variable for non-missing fy20 targets


## what about something more useful


## make it look a little nicer


## more useful still


```

`mutate` can be used to create a new column, or replace an exsisting column by keeping the same name
Below is a really good thing to add when working with the MSD

```{r}



```

We're going to tiptoe into condtions now. There is a lot to cover so we'll start slow.
A very simple but powerful function to use with `mutate` is `if_else()`. With `if_else()`, you first specify a logical statement, afterwards what needs to happen if the statement returns TRUE, and lastly what needs to happen if it’s FALSE. ([What is the difference between ifelse() and if_else?](https://medium.com/@HollyEmblem/chaining-if-else-statements-using-the-superheroes-dataset-with-dplyr-b13121777b4))

```{r}

# let's make some groups, shall we?



## note it dropped dedups
## usaid vs others



## make it a little nicer


## let's try an example with a logical operator
## how many mechs had targets for TX_CURR for USAID in 2019?

df %>% distinct(mech_name, fundingagency, fiscal_year)
## not helpful



```
