---
title: "Intro to R Markdown"
author: "Tim Essam | SI"
date: "6/9/2022"
output: html_document
params:
  indicator: HTS_TST
  period: FY50
  period_type: cumulative
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_chunk$set(echo = params$show_code)
```
#### This is a level four header - This is going to provide me a summary of the parameters used

```{r}
# Summary of parameters used in the document
glue::glue("Parameters used {params$indicator}, {params$period}, {params$period_type}")
```

Presentation recording: https://drive.google.com/file/d/1bICxJN_7x2ZaMT8xDCCB6rOTV5nwjBit/view


## This is a level two header
Here is where I describe the r chunk below. This chunk will load the libraries we need to make a dynamic document.

```{r load packages and testing data, message = FALSE, warning = FALSE, echo = TRUE}
library(glitr)
library(glamr)
library(gophr)
library(tidyverse)
library(tidytext)
library(extrafont)
library(glue)
library(scales)
library(gt)
```

### This is a level three header
Below, we load the `hts` dataset and filter it based on the parameters we defined in the YAML chunk at the top. 
Then, we create a ggplot object based on the new data frame. You could also use parameters in the ggplot call to produce a set of plots.

```{r}
hts_viz <- hts %>% 
  filter(indicator == params$indicator, period_type == params$period_type, period == params$period) %>% 
  mutate(mod_order = reorder_within(modality, value, prime_partner_name))

hts_viz %>% 
  ggplot(aes(x = value, y = mod_order)) + 
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~prime_partner_name, scales = "free_y") +
  si_style(facet_space = 0.5, text_scale = 0.65) +
  scale_x_continuous(labels = comma) +
  labs(title = glue("TESTING {str_to_upper(params$period_type)} FOR {params$indicator} IN {params$period}"),
       caption = glue("Source: glitr::hts dataset; Created on {Sys.Date()}"))
```


## Including Tables

You can also embed tables a document. For example, we can use `gt` to create a summary table. 

```{r pressure, echo=FALSE}
hts_viz %>% 
  select(-mod_order) %>% 
  pivot_wider(names_from = modality) %>% 
  gt(groupname_col = "indicator") %>% 
  cols_hide(columns = c(1, 3, 4, 5)) %>% 
  tab_options(table.font.size = 10) %>% 
  sub_missing(columns = everything(), missing_text = "-") %>% 
  fmt_number(columns = where(is.numeric), sep_mark = ",", decimals = 0) %>% 
  tab_header(title = glue("{params$period_type} for {params$indicator} in {params$period}") %>% str_to_upper)
  
```
#### Dynamic text based on R code
You can use R code inside the markdown chunk to generate a dynamic value. Below we nest the code `hts_viz %>% filter(modality == "Index") %>% summarize(total = comma(sum(value)))` to generate a dynamic result for Index totals.

The total number of Index test ``r params$period_type`` in `r params$period` for `r params$indicator` is `r hts_viz %>% filter(modality == "Index") %>% summarize(total = comma(sum(value)))`
