---
title: "ggplot + MER"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

Today we will be diving into the some exploratory analysis with R, getting familiar with visualizing data using `ggplot2` which is part of the [tidyverse](https://www.tidyverse.org/).

To get started, you'll need to install a few different packages to work in R. Unlike proprietary software like Excel, Stata, SPSS, etc. where all the functions are built into shipped software, R is open source and allows on contributers to develop and share useful code in "packages." 

The main way to install software in R is to use `install.package()`. This function (built into R) allows you access and install software from [CRAN](https://cran.r-project.org/). 


```{r install}
install.packages("tidyverse")
install.packages("devtools")
install.packages("skimr")
install.packages("scales")
install.packages("extrafont")
install.packages("patchwork")
```

Users can also access additional packages not available on CRAN for various reasons (still in development, small scale, etc). We will also be using a package developed for use with the PEPFAR MSD, called [ICPIutilities](https://github.com/ICPI/ICPIutilities/blob/master/README.md) and is able to be installed from GitHub (via the `devtools` package).

```{r install_github}
devtools::install_github("ICPI/ICPIutilities")
```

Let's take a look at all the packages loaded with `install.packages("tidyverse")`, which actually installs a host of packages for you to use. We can see what those are through running the code below.

```{r tidyverse}
tidyverse::tidyverse_packages()
```

Great. We have the options to utitlize these packages now that they are installed,  Every time we start up a new R session, we have load the packages we want to use, using the function, `library()`.

```{r load, echo=FALSE}
library("tidyverse")
library("devtools")
library("skimr")
library("scales")
library("extrafont")
library("patchwork")
```


## Loading the data

You can use lots of different types of data in R like csv, txt, and json or even proprietary formats like xlsx and .dta. 

Today we'll be using a csv file from our training dataset that has been filtered down to a manageable size, focusing on one indicator in one operating unit.

To load the data, we will rely on a packge from the tidyverse called `readr`. One of the cool features is we can actually load data from a url and we don't need it stored locally (although, please don't store any real PEPFAR data publically).

Today's dataset is available on [Github](https://github.com/USAID-OHA-SI/coRps/blob/master/2020-03-30/FY20_MilkyWay_Testing.csv). Let's load the data locally.

```{r}
url <- "https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-03-30/FY20_MilkyWay_Testing.csv"

df_hts <- read_csv(url)


```

Last session we went over some of the data exploration after loading the data and spent most of our time on the aesthetics in ggplot structured based on R for Data Science.


## FACETS

```{r}
ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality)) +
  facet_wrap(~operatingunit, scales = "free")
 
ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality)) +
  facet_wrap(~fct_lump(modality, 3))

df_hts %>% 
  filter(is_index == TRUE) %>% 
  ggplot() +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality)) +
  facet_grid(modality ~ operatingunit)
  
```


## Geometic Objects

```{r}
ggplot(df_hts) +
  geom_smooth(aes(HTS_TST_POS, Positivity), method = "lm", formula = y~x, se = FALSE) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST))


ggplot(df_hts, aes(HTS_TST_POS, Positivity)) +
  geom_point()
  
```

## Transformations

```{r}
ggplot(df_hts, aes(operatingunit)) +
  geom_bar()


ggplot(df_hts, aes(operatingunit, HTS_TST_POS)) +
  geom_bar(stat = "identity")

ggplot(df_hts, aes(operatingunit, HTS_TST_POS)) +
  geom_col()


ggplot(df_hts, aes(operatingunit, HTS_TST_POS, fill = modality)) +
  geom_col()


ggplot(df_hts, aes(operatingunit, HTS_TST_POS, fill = modality)) +
  geom_col() +
  facet_wrap(. ~ modality, scales = "free_x")

ggplot(df_hts, aes(operatingunit, HTS_TST_POS, fill = modality)) +
  geom_col(position = "dodge")

ggplot(df_hts, aes(Positivity, operatingunit)) +
         geom_point(size = 6, alpha = .5,
                    position = "jitter")

ggplot(df_hts, aes(Positivity, operatingunit)) +
         geom_jitter(size = 6, alpha = .5,
                    height = .15)

ggplot(df_hts) +
  geom_col(aes(modality, HTS_TST)) +
  coord_flip()

```



