---
title: "ggplot Exercises | 2020-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overivew

Below are three questions to work on to test our ggplot skills.


## Setup

```{r}
#Dependencies
  library(tidyverse)
  library(scales)

#import data
  df_trends <- read_csv("https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-04-13/FY18-20_Jupiter_txtrends.csv")
  
  df_linkage <- read_csv("https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-04-13/FY19_Saturn_linkage.csv")
  
  df_achievement <- read_csv("https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-04-13/FY20_Neptune_txnew_achievement.csv")
```


## Linkage

The USAID/Saturn team is interested in knowing where partners are seeing less than ideal linkage to treatment. Using `df_linkage`, help them create a scatter plot that looks at testing and treatment, flagging where they're far from a 1:1 relationship. Consider using different aesthetics to convey a message to your audience.


```{r linkage}

glimpse(df_linkage)

ggplot(df_linkage, aes(HTS_TST_POS, TX_NEW)) +
  geom_point(aes(colour = linkage), size = 6, lwd = .5, alpha = .5) +
  geom_abline(intercept = 0, slope = 1, size = 1, color = 'red') +
  theme_minimal()





```


## Treatment Trends

A number of questions have come up recently about treatment trends in Jupiter. You will want to use `df_trends` to review trends in TX_CURR, TX_NEW, and TX_NET_NEW from FY18 to the present visualize through a line chart. Consider using different aesthetics to convey a message to your audience.



```{r trends}

glimpse(df_trends)

ggplot(df_trends, aes(period, val, fill = indicator)) +
  geom_col(show.legend = F) +
  facet_wrap(~indicator, nrow = 3) +
  coord_flip() +
  theme_minimal() 

df_trends %>% 
  filter(indicator == 'TX_NET_NEW') %>% 
  ggplot(aes(period, val, fill = indicator)) +
    geom_col(show.legend = F) +
    coord_polar() +
    theme_minimal() 


```

## Target Achievement

In preparation for the FY20Q1 POART, you want to assess different partners' achievement for TX_NEW by sex in Neptune. You will want to show both the real results/targets as well as the percent achievement. Consider using different aesthetics to convey a message to your audience.


```{r achievement}

glimpse(df_achievement)

ggplot(df_achievement)





```

