---
title: Data Transformation
date: 2020-05-29
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

You have a dataset of Jupiter's FY20Q1 TX_CURR results and a separate dataset of Jupiter's FY20 TX_CURR targets. You have been asked to provide a list of the mechanisms that have targets but did not report results.

Exercise data can be found here:

results: https://raw.githubusercontent.com/USAID-OHA-SI/coRps/gina/2020-05-29/Jupiter_TX_FY20_Results.txt

targets: https://raw.githubusercontent.com/USAID-OHA-SI/coRps/gina/2020-05-29/Jupiter_TX_FY20_Targets.txt


## Load required libraries

Load R Packages: tidyverse, readxl, ICPIutilities

```{r}
library(...)
library(...)
library(...)
```


Import data - which readr function do you use for the text flie?

```{r}

df_results <-read_tsv("https://raw.githubusercontent.com/USAID-OHA-SI/coRps/gina/2020-05-29/Jupiter_TX_FY20_Results.txt")

```

```{r}

df_targets <- read_tsv("https://raw.githubusercontent.com/USAID-OHA-SI/coRps/gina/2020-05-29/Jupiter_TX_FY20_Targets.txt")
```

Decide which join you should use to get a list of mechanisms with targets, but no results

```{r}


```

You have been asked to provide the full Jupiter TX_CURR results and targets in a single file. How would you do this?

```{r}
#append results & targets


```

