---
title: "FY23 OHA Annual Top Talking Points - Snippet"
output: html_document
date: "3/18/2024"

---
```{r option, include=FALSE, echo=FALSE}
options(scipen=999)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

  library(tidyverse)
  library(glitr)
  library(glamr)
  library(gophr)
  library(systemfonts)
  library(scales)
  library(tidytext)
  library(patchwork)
  library(ggtext)
  library(glue)
  library(janitor)
  library(lubridate)
  library(googlesheets4)
  library(mindthegap)

load_secrets()
  
```
# Purpose

Last Updated: 3/18/2024

Previously used for: Data for World AIDS Day 2022, 2022 OHA Top Talking Points, PPR

Audience: The public, Agency stakeholders, partners, HIV/AIDS stakeholders/activists, the development community

Purpose: The purpose of this document is to create one place for the data points that can be shared by OHA leadership at events, in presentations, etc. 

```{r data, include=FALSE, echo=FALSE}
#Current MSD
filepath <- si_path() %>% 
  return_latest("MER_Structured_Datasets_OU_IM_FY21-24")

# Grab metadata
get_metadata(filepath) 

# MSD PREP --------------------------------------

#read MSD
df_msd <- read_psd(filepath)

#read NGA genie file
df_nga <- si_path() %>% 
  return_latest("All indicators_FY23Q4_Nigeria") %>% 
  read_csv()

#select only names in MSD
names_to_keep <- names(df_msd)
setdiff(names(df_msd), names(df_nga)) #only qtrs
setdiff(names(df_nga), names(df_msd))

#select all names from NGA file that match MSD (minus qtr1-qtr4)
df_nga_new <- df_nga %>% 
  select(any_of(names_to_keep))

#rbind together removing TZ and NGA for FY23 ONLY
df <- df_msd %>% 
  filter(!(operatingunit=="Nigeria" & fiscal_year=="2023"),
         !(operatingunit=="Tanzania" & fiscal_year=="2023")) %>% 
  select(-c(qtr1:qtr4)) %>% 
  rbind(df_nga_new)

#Q4 TZA Daily Genie 
df_TZA <- si_path() %>%
  return_latest("Genie-OUByIMs-Tanzania-Daily-2024-02-07") %>%
  read_psd()

#exclude extra columns 
df_msd_TZA <- df_TZA %>% 
  select(-c(qtr1:qtr4)) 

#bind TZA data to clean MSD w/ NGA file
df_final <- df %>% 
  rbind(df_msd_TZA)

# load source functions
#source("Scripts/archive/99_utilities.R")

```

```{r data_prep_functions, include=FALSE, echo=FALSE}

#LOCAL PARTNER PREP
#Read in the google sheet hyperfile with local partner
sheet_id <- "1MQviknJkJDttGdNEJeNaYPKmHCw6BqPuJ0C5cslV5IE"

df_partner <- read_sheet(sheet_id, sheet = "MechID-PartnerType", range = "A:B") %>% 
  clean_names() %>% 
  rename(mech_code = mechanism_id) %>% 
  mutate(mech_code = as.character(mech_code),
         partner_type = case_when(partner_type == "Regional" ~ "Local",
                                  partner_type == "TBD Local" ~ "Local", TRUE ~ partner_type))

#FUNCTIONS----------------------------------------------------------------------

#clean number function
clean_number <- function(x, digits = 0){
  dplyr::case_when(x >= 1e9 ~ glue("{round(x/1e9, digits)} billion"),
                   x >= 1e6 ~ glue("{round(x/1e6, digits)} million"),
                   x >= 1e3 ~ glue("{round(x/1e3, digits)} thousand"),
                   TRUE ~ glue("{x}"))
}

#LP Function - grabs local partner share of results
#result_type = value or share
#(clean = TRUE formats the value)
grab_lp_results <- function(indic, result_type, clean = FALSE) {
  
  df_munge <- suppressMessages(df_final %>% 
                                 # bind_rows(df_arch) %>%
                                 left_join(df_partner, by = c("mech_code")) %>%
                                 filter(funding_agency == "USAID",
                                        indicator == indic,
                                        standardizeddisaggregate == "Total Numerator") %>% 
                                 group_by(funding_agency, fiscal_year, indicator, partner_type) %>% 
                                 summarise(across(cumulative, sum, na.rm = TRUE)) %>%
                                 ungroup() %>% 
                                 filter(partner_type != "TBD") %>%
                                 #filter(fiscal_year != 2022) %>%
                                 pivot_wider(names_from = partner_type, values_from = cumulative) %>%
                                 group_by(fiscal_year) %>%
                                 mutate(Total = International + Local,
                                        share = Local / Total)  %>%
                                 pivot_longer(cols = International:Total, names_to = "partner_type"))
  
  title_info_lp <- df_munge %>% 
    filter(partner_type == "Local", fiscal_year == metadata$curr_fy) %>% 
    select(fiscal_year, indicator, value, share) %>% 
    mutate(
      value = value,#change to 1 if you want 1 decimal accuracy
      share = percent(round(share, 2))) %>% 
    pull(result_type)
  
  if (clean ==  TRUE) {
    title_info_lp <- df_munge %>% 
      filter(partner_type == "Local", fiscal_year == metadata$curr_fy) %>% 
      select(fiscal_year, indicator, value, share) %>% 
      mutate(
        value = value %>% clean_number(1),#change to 1 if you want 1 decimal accuracy
        share = percent(round(share, 2))) %>% 
      pull(result_type)
  }
    
    return(title_info_lp)
}


# pull KP share of MER results (clean = TRUE formats the value)
pull_kp <- function(indic, clean = FALSE) {
  
  kp_val <- df_final %>%
    filter(funding_agency == "USAID",
           str_detect(standardizeddisaggregate, "KeyPop(?!\\/Status)"),
           indicator %in% c("PrEP_NEW", "TX_CURR", "KP_PREV"),
           fiscal_year == metadata$curr_fy) %>%
    count(indicator, wt = cumulative) 
  

  
  if (clean == TRUE) {
   kp_val <- kp_val %>% 
    mutate(n = clean_number(n, 1)) %>% 
    pivot_wider(names_from = indicator, values_from = "n") %>% 
    pull(indic)  
  } else {
    
    kp_val <- kp_val %>% 
      pivot_wider(names_from = indicator, values_from = "n") %>% 
      pull(indic)  
  }
  
  return(kp_val)
}

load_secrets()
```


```{r prevention_prep, echo=FALSE, include = FALSE}

prep_val <- df_final %>%
  filter(funding_agency == "USAID",
         standardizeddisaggregate %in% c("Total Numerator"),
         indicator %in% c("PrEP_NEW"),
         fiscal_year == metadata$curr_fy) %>%
  summarise(across(starts_with("cumulative"), sum, na.rm = T)) %>% 
  mutate(cumulative = clean_number(cumulative)) %>% 
  pull(cumulative)

prep_val_country <- df_final %>% 
  filter(funding_agency == "USAID",
         standardizeddisaggregate %in% c("Total Numerator"),
         indicator %in% c("PrEP_NEW"),
         fiscal_year == metadata$curr_fy) %>%
  distinct(country) %>% nrow()

lp_prep_val_clean <- grab_lp_results("PrEP_NEW", "value", clean = TRUE) 
lp_prep_val <- grab_lp_results("PrEP_NEW", "value", clean = FALSE) 
lp_prep_share <- grab_lp_results("PrEP_NEW", "share")

#kp prep
kp_prep_val_clean <- pull_kp("PrEP_NEW", clean = TRUE)
kp_prep_val <- pull_kp("PrEP_NEW", clean = FALSE)

#AGYW prep
prep_agyw_val <- df_final %>%
  filter(funding_agency == "USAID",
         standardizeddisaggregate %in% c("Age/Sex"),
         indicator %in% c("PrEP_NEW"),
         fiscal_year == metadata$curr_fy,
         sex == "Female",
         age_2019 %in% c("10-14","15-19", "20-24")) %>%
  summarise(across(starts_with("cumulative"), sum, na.rm = T)) %>% 
  pull(cumulative)

prep_agyw_val_clean <- df_final %>%
  filter(funding_agency == "USAID",
         standardizeddisaggregate %in% c("Age/Sex"),
         indicator %in% c("PrEP_NEW"),
         fiscal_year == metadata$curr_fy,
         sex == "Female",
         age_2019 %in% c("10-14","15-19", "20-24")) %>%
  summarise(across(starts_with("cumulative"), sum, na.rm = T)) %>% 
  mutate(cumulative = clean_number(cumulative, 0)) %>%
  pull(cumulative)
```

**FY23 OHA Top Talking Points**

**Prevention**

1. In 2023, USAID initiated pre-exposure prophylaxis (PrEP) for HIV prevention in over **`r prep_val`** people across **`r prep_val_country`** countries. USAID, with PEPFAR resources, supports the UNAIDS goal of having 10 million people on PrEP globally by 2025. 

2. USAID Local Partners initiated oral PrEP in almost **`r lp_prep_val_clean`** [**`r lp_prep_val`**] individuals in 2023, representing **`r lp_prep_share`** of all USAID PrEP initiations. 

3. USAID initiated PrEP in more than **`r kp_prep_val_clean`** [**`r kp_prep_val`**] members of key populations (including sex workers, transgender people, men who have sex with men, people in prisons, and people who inject drugs) and more than **`r prep_agyw_val_clean`** [**`r prep_agyw_val`**] adolescent girls and young women (AGYW) aged 15-24 years in 2023. Together, key populations (KP) and AGYW constitute more than half of all USAID PrEP initiations.