---
title: "Working with data in JSON Format"
output: html_document
author: "B.Kagniniwa"
organisation: "USAID"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plan

1. Overview - What is JSON (GeoJSON)?
2. How to read JSON (GeoJSON) data
3. DataFrame to JSON
4. Resources


## Overview

JSON stands for JavaScript Object Notation and was initial conceived 
as a way to transport JavaScript object between a browser and a web server.

JS = JavaScript (Initial used in JavaSprit Applications)
O  = Object 
N  = Notation (A way to represent data)

JSON was discovered and popularized by [Douglas Crockford](https://www.crockford.com/about.html) in the 2000s,
and then standardized in 2013 as a Lightweight Data-interchange Format.

Today, JSON is the most polular data exchange format and is used by almost all programming languages. JSON has surpassed XML, CSV formats.

JSON is built on 2 main structures:

* JSON Object, a collection of name/value pairs
  + `JavaScript` => `object`, `Python` => `dictionary`, `R` => `named list`
* JSON Array, a ordered list of values
  + `JavaScript` => `array`, `Python` => `list`, `R` => `vector`, `list`
  
### Examples of JSON data:

`["mongo", 5, true, 1.95, null]`

`{"group":"fruit", "name":"mongo", "price":.99, "sale":true, "total":7}`

`{"group":"fruit", "name":"mongo", "price": {"normal": 1.99, "sale":false}}`

`{"group":"fruit", "name":"mongo", "price": [1.99, false]}`

`{
    "level": 7,
    "name": "Achikondi Women Community Friendly Services Clinic",
    "id": "FhQhjdiZXxo",
    "geometry": {
        "type": "Point",
        "coordinates": [33.7793, -13.95473]
    }
}`

### Examples of JSON Resources:

[DATIM API Resources](https://www.datim.org/api/resources.json)

[DATIM Organisation Units](https://www.datim.org/api/organisationUnits.json)

[Malawi Org hierarchy data](https://final.datim.org/api/organisationUnits?filter=path:like:lZsCb6y0KDX&fields=id,name,path,level,geometry&paging=false&format=json)


## 2. Read data in JSON Format

Load required R Packages

```{r}
library(tidyverse) # Tablular Data wrangling
library(jsonlite)  # JSON data formating
library(janitor)   # Helper package
```

Let's read in some dummy raw json data

```{r}
# JSON Array
raw_json0 <- '["mongo", 5, true, 1.95, null]'

print(raw_json0)

# JSON Complex Objects
raw_json1 <- '[{"group":"fruit", "name":"mongo", "price": {"actual": 1.99, "sale":false, "rate": 0, "stores": [22, 890]}}]'

print(raw_json1)

raw_json2 <- '[{"level": 7,"name": "Achikondi Women Community Friendly Services Clinic","id": "FhQhjdiZXxo","geometry": {"type": "Point","coordinates": [33.7793, -13.95473]}}]'

print(raw_json2)

```

These look like a text!

JSON is a text based data exchange format, so we will need to transform it into something that R can understand.

`fromJSON` function from `jsonlite` package will be helpfull in this case.


```{r}

# jsonlite::fromJSON
dta_json1 <- raw_json1 %>% fromJSON()

class(dta_json1)

dta_json1

```


This is better than a text. What can you do with a df nested in another df?
Take a look at the price column? Can you tell what's inside?


```{r}

# Convert json list to df
dta_json1 <- raw_json1 %>% 
  fromJSON() %>% 
  as_tibble()

class(dta_json1)

dta_json1

```

About the same format. How can we extract the nested df?

`as_tibble()` uses the main name/key level of the object as variable names.
How do we get to the next level? Let's try to flatten the JSON Object.


```{r}
# For fromJSON to flatten the object
dta_json1 <- raw_json1 %>% 
  fromJSON(flatten = TRUE) %>% 
  as_tibble()

dta_json1

```


Yeaah, making progress. Really? The `price.stores` is still a list!
What can we unpack this column?

```{r}

# Unnest list longer
dta_json1 <- raw_json1 %>% 
  fromJSON(flatten = TRUE) %>% 
  unnest_longer("price.stores")

dta_json1

# Unnest lists wider
dta_json1 <- raw_json1 %>% 
  fromJSON(flatten = TRUE) %>% 
  unnest_wider("price.stores", names_sep = "_")

dta_json1

```


Great. This version is better. Let's clean up the column names


```{r}

# Replace dots with _
dta_json1 <- dta_json1 %>% 
  clean_names()

dta_json1

```


Let's apply these to the second JSON object


```{r}

# Raw json data
# {
#     "level": 7,
#     "name": "Achikondi Women Community Friendly Services Clinic",
#     "id": "FhQhjdiZXxo",
#     "geometry": {
#         "type": "Point",
#         "coordinates": [33.7793, -13.95473]
#     }
# }

dta_json2 <- raw_json2 %>% 
  fromJSON(flatten = TRUE) %>% 
  unnest_wider("geometry.coordinates", names_sep = "_") %>% 
  clean_names() %>% 
  rename(
    longitude = geometry_coordinates_1,
    latitude = geometry_coordinates_2
  )

dta_json2 %>% glimpse()

dta_json2

```

## 3. JSON to R DataFron

```{r}

toJSON(dta_json2)

```

## 4. Resources

[JSON](https://www.json.org/json-en.html)

[W3SCOOLS](https://www.w3schools.com/js/js_json_intro.asp)

[JS Beautifier - Parser](https://beautifier.io/)

[GeoJSON](https://geojson.org/)
