---
title: "beautiful_tables_with_gt"
author: "jdavis"
date: "1/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Today we're going to be learning how to create presentation quality tables in R using the `gt` package. 


The basic outline is as follows:
  *munging your data into the tabluar structure you want with `tidyverse`
  *converting your table to a `gt` object
  *reviewing + formatting the parts of your table
    *formatting values
      *title
      *sub title
    *conditional formatting
  *summary rows
  *saving/exporting your work

  
## Libraries + globals
  
```{r}
library(tidyverse)
library(glitr)
#install.packages("gt")
library(gt)
library(googledrive)
library(googlesheets4)
library(ICPIutilities)

```
  
##munging your data
  We're going to look at a simple table that's in long format and convert to wide using `pivot_wider
  
```{r}
## lines 47 - 60 have been done for you
# raw <- read_msd("your_ou_by_IM_MSD")
# # 
#   filter(operatingunit == "Zambia",
#          standardizeddisaggregate == "Total Numerator",
#          indicator %in% c("HTS_TST", "HTS_TST_POS", "TX_NEW", "TX_CURR", "TX_NET_NEW", "VMMC_CIRC", "PrEP_NEW"),
#          fiscal_year %in% c(2020, 2021))%>% 
#   reshape_msd("long") %>% 
#   filter(period != "fy2021q1")
# # 
# df <- df %>% 
#   dplyr::group_by(indicator, period) %>% 
#   dplyr::summarise(value = sum(val))

## have a look at `gt`

ls("package:gt")

##read in data

df <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1iiwaESC45G9waP7CxQLEbqo1Qq6vw-VxZAUh1qSS5yg/edit#gid=0",
                                sheet = "Sheet1")
#have a look
glimpse(df)


df <- df %>%
  filter(fundingagency == "USAID") %>% 
  tidyr::pivot_wider(names_from = period,
              values_from = value)

glimpse(df)
  
  df <- df %>%
    dplyr::relocate(fy2020cumulative, .before = fy2021_targets) %>%
    dplyr::relocate(fy2020_targets, .after = fy2020cumulative) %>% 
    mutate(indicator = fct_relevel(indicator, "HTS_TST",
                                   "HTS_TST_POS",
                                   "TX_NEW",
                                   "TX_CURR",
                                   "TX_NET_NEW",
                                   "VMMC_CIRC",
                                   "PrEP_NEW")) %>% 
    arrange(indicator) %>% 
    rename(`FY20 Q1` = fy2020q1,
      `FY20 Q2` = fy2020q2,
      `FY20 Q3` = fy2020q3,
      `FY20 Q4` = fy2020q4,
      `FY20 Total`  = fy2020cumulative,
      `Fy20 Targets` = fy2020_targets,
      `FY21 Targets` = fy2021_targets)
  
glimpse(df)

```

## converting your table to a `gt` object

Now that your data is the form that you want, we're going to pass to `gt`. You have to do all your munging before you get to this
point since you cannot call `dplyr` from `gt`. `gt` converts your dataframe to a list and then allows you to format the objects
within that list. This can be a bit of a jarring thing to deal with; as social scientists and not computer science people
we are more comfortable dealing with dataframes (rows and columns) than lists. But lists are cool, don't worry.

```{r}

df %>% 
  gt()

## you can see that without doing any modification whatsoever, `gt` spits out a pretty good table with just the defaults
## but we want to clean this up a bit

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator")
```

 
Note that we included "fundingagency" as the `groupname_col` even tho it's not displayed (we filtered on it earlier) tho it is
technically the grouping var. If we had left CDC in there for ex, you would see two groups

`gt` uses what you pass to it in `groupname_col` for other operations, so if you are finding that something is not working, try
adding something the `groupname_col`. You can also change the order of the groups with `row_group_order`

##reviewing + formatting the parts of your table
  
There are many different ways you can format you table in `gt`, we're going to focus on the `fmt*`
`fmt*` has a ton of options, pretty much whatever you would think to need, they have.


```{r}

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0
  )

# a few things to note: become familiar with the defaults for some of these (see help) for ex, we didn't include 
# `use_seps = TRUE` after line 136. This is true by default and defaults to a "," separator, but you can change this

```

Getting there, but lets get rid of those NA values.

```{r}

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0
  ) %>% 
fmt_missing(columns = everything(),
            missing_text = "-")

````






























