---
title: "beautiful_tables_with_gt"
author: "jdavis"
date: "1/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Today we're going to be learning how to create presentation quality tables in R using the `gt` package. 
We'll be reproducing the table located here:
https://docs.google.com/presentation/d/1tR3a81fd5mtkAX-RkX3hZv7skahkHryIrbrI7mli1Pc/edit#slide=id.p6


The basic outline is as follows:
  *munging your data into the tabluar structure you want with `tidyverse`
  *converting your table to a `gt` object
  *reviewing + formatting the parts of your table
    *formatting values
      *title
      *sub title
    *conditional formatting
  *summary rows
  *saving/exporting your work

  
## Libraries + globals
  
```{r}
library(tidyverse)
library(glamr)
#install.packages("gt")
library(gt)
library(googledrive)
library(googlesheets4)
library(ICPIutilities)
library(glitr)
library(RColorBrewer)
library(scales)
library(here)


pal <- RColorBrewer::brewer.pal(5, "Spectral")[2:5] ## we'll come back to this
save_path <- "2021-01-18"

```
  
##munging your data
  We're going to look at a simple table that's in long format and convert to wide using `pivot_wider
  
```{r}
## lines 57 - 68 have been done for you
# raw <- read_msd(here(folderpath_msd, "filename"))
# # 
#   filter(operatingunit == "Zambia",
#          standardizeddisaggregate == "Total Numerator",
#          indicator %in% c("HTS_TST", "HTS_TST_POS", "TX_NEW", "TX_CURR", "TX_NET_NEW", "VMMC_CIRC", "PrEP_NEW"),
#          fiscal_year %in% c(2020, 2021))%>% 
#   reshape_msd("long") %>% 
#   filter(period != "fy2021q1")
# # 
# df <- df %>% 
#   dplyr::group_by(indicator, period) %>% 
#   dplyr::summarise(value = sum(val))

## have a look at `gt`

ls("package:gt")

##read in data

gs4_auth() #or glamr::load_secrets() https://github.com/USAID-OHA-SI/glamr/blob/master/vignettes/credential-management.Rmd


df_raw <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1iiwaESC45G9waP7CxQLEbqo1Qq6vw-VxZAUh1qSS5yg/edit#gid=0",


#have a look
glimpse(df_raw)


df <- df_raw %>%
  filter(fundingagency == "USAID") %>% 
  tidyr::pivot_wider(names_from = period,
              values_from = value) %>% 
   mutate(`FY20 Achieved` =  fy2020cumulative/fy2020_targets) %>% 
  rename(`FY20 Q1` = fy2020q1,
      `FY20 Q2` = fy2020q2,
      `FY20 Q3` = fy2020q3,
      `FY20 Q4` = fy2020q4,
      `FY20 Total`  = fy2020cumulative,
      `Fy20 Targets` = fy2020_targets,
      `FY21 Targets` = fy2021_targets) %>% 
    mutate(`FY20 Achieved` =  `FY20 Total`/`Fy20 Targets`)

glimpse(df)
  
  df <- df %>%
    dplyr::relocate(`FY20 Total`, .before = `FY21 Targets`) %>% 
    dplyr::relocate(`FY20 Achieved`, .before = `FY21 Targets`) %>%
    dplyr::relocate(`Fy20 Targets`, .after = `FY20 Total`) %>% 
    mutate(indicator = fct_relevel(indicator, "HTS_TST",
                                   "HTS_TST_POS",
                                   "TX_NEW",
                                   "TX_CURR",
                                   "TX_NET_NEW",
                                   "VMMC_CIRC",
                                   "PrEP_NEW")) %>% 
    arrange(indicator)
  
glimpse(df)

```

## converting your table to a `gt` object

Now that your data is the form that you want, we're going to pass to `gt`. You have to do all your munging before you get to this
point since you cannot call `dplyr` from `gt`. `gt` converts your dataframe to a list and then allows you to format the objects
within that list. This can be a bit of a jarring thing to deal with; as social scientists and not computer science people
we are more comfortable dealing with dataframes (rows and columns) than lists. But lists are cool, don't worry.

One thing to remember is that if you `tbl_*` in a function, it's refering to the part of the table that function is modifying
You can see the parts of the table visually here https://gt.rstudio.com/articles/intro-creating-gt-tables.html

```{r}

df %>% 
  gt()

## you can see that without doing any modification whatsoever, `gt` spits out a pretty good table with just the defaults
## but we want to clean this up a bit

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator")
```

 
Note that we included "fundingagency" as the `groupname_col` even tho it's not displayed (we filtered on it earlier) tho it is
technically the grouping var. If we had left CDC in there for ex, you would see two groups

`gt` uses what you pass to it in `groupname_col` for other operations, so if you are finding that something is not working, try
adding something the `groupname_col`. You can also change the order of the groups with `row_group_order`

##reviewing + formatting the parts of your table
  
There are many different ways you can format you table in `gt`, we're going to focus on the `fmt*`
`fmt*` has a ton of options, pretty much whatever you would think to need, they have.


```{r}

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0
  ) %>% 
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0
  )

# a few things to note: become familiar with the defaults for some of these (see help) for ex, we didn't include 
# `use_seps = TRUE` after line 136. This is true by default and defaults to a "," separator, but you can change this

```

Getting there, but lets get rid of those NA values.

```{r}

df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0
  ) %>%
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0
  ) %>% 
fmt_missing(columns = everything(),
            missing_text = "-")

```


Ok basically there, but lets adjust the formatting a bit before we move into the conditional formatting
To do that, we're going to the `tab_options` which lets you modify the characteristics of the whole table,
rather than parts (which we cover in a bit). The two below are `table.font.names` and `data_row.padding`

We're also going to add `cols_witdth` which allows you to play with the size (spacing) of various columns
The `cols_*` functions in `gt` have a number of `dplyr`-like functionalities that let you move things around
(similar to `arrange`) so if you are deep in a `gt` workflow and it would be easier than modding your original
df, consider these.


```{r}
df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0) %>%
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0) %>% 
fmt_missing(columns = everything(),
            missing_text = "-") %>% 
  tab_options(
    table.font.names = "Source Sans Pro"
    ) %>% 
   cols_width(
    vars(indicator) ~ px(140),
    everything() ~ px(80)
  )
  

```

Next we're going to look at `tab_styles` which is a real powerhouse in this package. It basically lets you do anything
to any cell, row, or columns. The two basic components are the style (what are you doing) and the location (where are you doing it)

We'll start by adding some cell borders


```{r}
df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0) %>%
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0) %>% 
fmt_missing(columns = everything(),
            missing_text = "-") %>% 
  tab_options(
    table.font.names = "Source Sans Pro") %>%
    cols_width(
    vars(indicator) ~ px(140),
    everything() ~ px(80)
  ) %>% 
  tab_style(
    style = cell_borders(
      sides = "right",
      weight = px(1.5),
    ),
    locations = cells_body(
      columns = everything(),
      rows = everything()
    ))

```

So now that we have our table in the basic format we want, let's add the shading based on value which is analogous to conditional
formatting

```{r}
df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0) %>%
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0) %>% 
fmt_missing(columns = everything(),
            missing_text = "-") %>% 
  tab_options(
    table.font.names = "Source Sans Pro") %>%
    cols_width(
    vars(indicator) ~ px(140),
    everything() ~ px(80)
  ) %>% 
  tab_style(
    style = cell_borders(
      sides = "right",
      weight = px(1.5),
    ),
    locations = cells_body(
      columns = everything(),
      rows = everything()
    )) %>% 
    tab_style(style = cell_fill(color = pal[4]),      ## defining the what (the 4th value of the pal object)
              locations = cells_body(                 ## telling it where (ie, the body of a cell)
                columns = vars(`FY20 Achieved`),      ## which col this refers to (note `vars()`)
                rows = `FY20 Achieved` >= 1.1)) %>%   ## the argument
    tab_style(style = cell_fill(color = pal[3]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < 1.1)) %>% 
    tab_style(style = cell_fill(color = pal[2]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < .9)) %>% 
    tab_style(style = cell_fill(color = pal[1]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < .75))

```


Let's add the finishing touches
-Title (very important) and subtitle (maybe important)
-Source (never leave this out)
Title,subtitle, and footnote are very easy and both can take advantage of markdown if you need it by adding `md()`
`tab_header`
`tab_footnote`
`tab_source_note`

```{r}
tbl <- df %>% 
  gt(groupname_col = "fundingagency",
    rowname_col = "indicator") %>% 
  fmt_number(
    columns = vars(
      `FY20 Q1`,
      `FY20 Q2`,
      `FY20 Q3`,
      `FY20 Q4`,
      `FY20 Total`,
      `Fy20 Targets`,
      `FY21 Targets`),
    decimals = 0) %>%
  fmt_percent(
    columns = vars(`FY20 Achieved`),
    decimals = 0) %>% 
fmt_missing(columns = everything(),
            missing_text = "-") %>% 
  tab_options(
    table.font.names = "Source Sans Pro") %>%
    cols_width(
    vars(indicator) ~ px(140),
    everything() ~ px(80)
  ) %>% 
  tab_style(
    style = cell_borders(
      sides = "right",
      weight = px(1.5),
    ),
    locations = cells_body(
      columns = everything(),
      rows = everything()
    )) %>% 
    tab_style(style = cell_fill(color = pal[4]),      ## defining the what (the 4th value of the pal object)
              locations = cells_body(                 ## telling it where (ie, the body of a cell)
                columns = vars(`FY20 Achieved`),      ## which col this refers to (note `vars()`)
                rows = `FY20 Achieved` >= 1.1)) %>%   ## the argument
    tab_style(style = cell_fill(color = pal[3]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < 1.1)) %>% 
    tab_style(style = cell_fill(color = pal[2]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < .9)) %>% 
    tab_style(style = cell_fill(color = pal[1]),
              locations = cells_body(
                columns = vars(`FY20 Achieved`),
                rows = `FY20 Achieved` < .75)) %>% 
  tab_header(title = "USAID results, performance, and targets FY20-21",
             subtitle = md("Super Timmy's **magic** tabler-ater")) %>% 
  tab_source_note("Source: DATIM MSD FY20Q4 2020-12-18")
  


```

Let's save.
`gt` has a few options. For our uses, we'll focus on .png and .pdf. You'll need to install the `webshot` package, which has most of the options for saving, we'll focus on `expand` and `zoom`

```{r}

gtsave(tbl, here(save_path,"corps_table.pdf"))
gtsave(tbl, here(save_path,"corps_table.png"))

#let's play with zoom
gtsave(tbl, here(save_path,"corps_table.pdf"), zoom = 1)


```



Ok here is a quick example with some grouping

```{r}

df2 <- df_raw %>% 
  filter(fundingagency %in% c("USAID", "HHS/CDC")) %>% 
  tidyr::pivot_wider(names_from = period,
              values_from = value) %>% 
   mutate(`FY20 Achieved` =  fy2020cumulative/fy2020_targets) %>% 
  rename(`FY20 Q1` = fy2020q1,
      `FY20 Q2` = fy2020q2,
      `FY20 Q3` = fy2020q3,
      `FY20 Q4` = fy2020q4,
      `FY20 Total`  = fy2020cumulative,
      `Fy20 Targets` = fy2020_targets,
      `FY21 Targets` = fy2021_targets) %>% 
    mutate(`FY20 Achieved` =  `FY20 Total`/`Fy20 Targets`)

glimpse(df2)
  
  df2 <- df2 %>%
    dplyr::relocate(`FY20 Total`, .before = `FY21 Targets`) %>% 
    dplyr::relocate(`FY20 Achieved`, .before = `FY21 Targets`) %>%
    dplyr::relocate(`Fy20 Targets`, .after = `FY20 Total`) %>% 
    mutate(indicator = fct_relevel(indicator, "HTS_TST",
                                   "HTS_TST_POS",
                                   "TX_NEW",
                                   "TX_CURR",
                                   "TX_NET_NEW",
                                   "VMMC_CIRC",
                                   "PrEP_NEW")) %>% 
    arrange(indicator)
  
glimpse(df2)

## let's pass that to CDC and see what it does

df2 %>% 
  gt() 

df2 %>% 
  gt(groupname_col = "fundingagency") %>% 
  row_group_order(
    groups = c("USAID", "HHS/CDC")
  )

```


Next a quick example looking at some summary functions

```{r}

df3 <- df_raw

df3 %>%
  filter(period == "fy2020cumulative") %>% 
  pivot_wider(names_from = indicator,
              values_from = value) %>% 
  gt(groupname_col = "indicator",
     rowname_col = "period") %>% 
  summary_rows(columns = vars("HTS_TST", "HTS_TST_POS", "TX_NEW", "TX_CURR", "TX_NET_NEW", "VMMC_CIRC", "PrEP_NEW"),
    fns = list(Total = ~sum(., na.rm = TRUE)),
                     formatter = fmt_number,
                     decimals = 0)
  

```


Finally some `gt` resources

https://themockup.blog/
An inspo album

https://gt.rstudio.com/index.html
Main reference

https://yutannihilation.github.io/gt/
More

https://malco.io/2020/05/16/replicating-an-nyt-table-of-swedish-covid-deaths-with-gt/
even more

























