---
title: "the problem with duplicates"
author: "jdavis"
date: "7/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Problem statement
Today we're going to cover a few tips and trick to deal with how to explore your data for duplicates and how to deal with them

This is good practice when working with datasets that are new to you or you are unsure about the quality

This is less of a problem with something like the MSD/genie because it goes through an extensive cleaning process

We'll go through a few simple examples using LMIS data

```{r}
#products
arvs <- c("Abacavir 300 mg, Tablet, 60 Tabs",
  "Abacavir/Lamivudine 600/300 mg Dispersible Tablet, 30 Tabs",
  "Abacavir/Lamivudine 600/300 mg Tablet, 30 Tablets",
  "Abacavir/Lamivudine 600/300 mg Tablet, 30 Tabs",
  "Atazanavir/Ritonavir 300/100 mg Tablet, 30 Tablets",
  "Atazanavir/ritonavir 300/100 mg Tablet, 30 Tabs",
  "Darunavir 600 mg Tablet, 60 Tabs",
  "Dolutegravir 50 mg Tablet, 30 Tabs",
  "Dolutegravir 50 mg Tablet, 30 Tabs",
  "Dolutegravir/Lamivudine/Tenofovir DF 50/150/300 mg Tablet, 30 Tabs",
  "Dolutegravir/Lamivudine/Tenofovir DF 50/300/300 mg Tablet, 180 Tabs",
  "Dolutegravir/Lamivudine/Tenofovir DF 50/300/300 mg Tablet, 30 Tabs",
  "Dolutegravir/Lamivudine/Tenofovir DF 50/300/300 mg Tablet, 90 Tabs",
  "Efavirenz 600 mg Tablet, 30 Tablets",
  "Efavirenz 600 mg Tablet, 30 Tabs",
  "Efavirenz/Emtricitabine/Tenofovir DF 600/200/300 mg Tablet, 30 Tabs",
  "Efavirenz/Lamivudine/Tenofovir DF 400/300/300 mg Tablet, 30 Tabs",
  "Efavirenz/Lamivudine/Tenofovir DF 400/300/300 mg Tablet, 90 Tabs",
  "Efavirenz/Lamivudine/Tenofovir DF 600/300/300 mg Tablet, 180 Tabs",
  "Efavirenz/Lamivudine/Tenofovir DF 600/300/300 mg Tablet, 30 Tablets",
  "Efavirenz/Lamivudine/Tenofovir DF 600/300/300 mg Tablet, 30 Tabs",
  "Efavirenz/Lamivudine/Tenofovir DF 600/300/300 mg Tablet, 90 Tabs",
  "Emtricitabine/Tenofovir DF 200/300 mg Tablet, 30 Tabs",
  "Emtricitabine/Tenofovir DF 300/300 mg Tablet, 30 Tabs",
  "Etravirine 100 mg Tablet, 120 Tabs",
  "Lamivudine 150 mg Tablet, 60 Tabs",
  "Lamivudine/Tenofovir DF 300/300 mg Tablet, 30 Tablets",
  "Lamivudine/Tenofovir DF 300/300 mg Tablet, 30 Tabs",
  "Lamivudine/Zidovudine 150/300 mg Tablet, 60 Tablets",
  "Lamivudine/Zidovudine 150/300 mg Tablet, 60 Tabs",
  "Lopinavir/ritonavir 200/50 mg Tablet, 120 Tablets",
  "Lopinavir/ritonavir 200/50 mg Tablet, 120 Tabs",
  "Nevirapine/Lamivudine/Zidovudine 200/150/300 mg Tablet, 60 Tablets",
  "Nevirapine/Lamivudine/Zidovudine 200/150/300 mg Tablet, 60 Tabs",
  "Raltegravir 400 mg Tablet, 60 Tabs")

```


```{r}

library(tidyverse)
library(vroom)

df <- vroom::vroom("C:/Users/Josh/Documents/GitHub/bigfoot/Data/2020-04_SC-FACT_Data.csv") %>% 
  rename_all(~tolower(.))

glimpse(df)

```

## What identifies a unique observation?

-Depends on your data
-What is an indicator, what is an observation?
-Is this tidy?


```{r}
#tidy and subset

df_long <- df %>%
  filter(period == "2020-03",
         product %in% arvs) %>% 
  gather(indicator, value, colnames(select_if(., is.numeric)), na.rm = TRUE) %>% 
  select(-facilitycd, -datimcode, -facility_mapped, -source, -`datim facility`) %>% 
  mutate_at(vars(country, facility), ~tolower(.)) %>% 
  ungroup()

#explore!

glimpse(df_long)
view(df_long)

df_long %>% distinct(country)

df_long %>% distinct(product)


```


## So lets look at what we have
## want to see for combo of indicators what is a unique observation
## using `n()` means that a variable named n will be assigned the number of rows (think number of observations) in the summarized data.

```{r}

df_long  <- df_long %>%
  group_by(country, facility, period, product, indicator, value) %>% 
  mutate(n = n(),
         dup_flag = row_number())

#what just happened?
df_long %>% View()

#filter it to the area of interest
df_long %>% filter(n > 1) %>% View()
#110 duplicates

#remove duplicate obs
df_long_dedup <- df_long %>% filter(n ==1)

df_long_dedup %>% filter(n > 1) %>% View()

#remove flags
df_long_dedup <- df_long_dedup %>% 
  select(-n, -dup_flag)

# Check if our filter worked
dim(df_long)[1] - dim(df_long_dedup)[1]

#try a different grouping
df_long2  <- df_long %>%
  group_by(country, snl1, facility,  period, product, indicator, value) %>% 
  mutate(n = n(),
         dup_flag = row_number())


```
















