---
title: "Making Maps with gisr package"
author: "B.Kagniniwa | G.Sarfaty"
date: "2020-07-20"
output: html_document
---
  
## Making Maps with R

Introduction to Mapping and Geospatial Analytics packages: sp, raster, sf, rnaturalearth, and gisr

1) Install R Packages for Geospatial Analytics
2) Spatial Data Acquisition
3) Spatial Data Visualization
4) Integration of Spatial Dimension to HQ Data Reviews and other daily work


## Spatial Packages

Let's install couple R Packages useful for Mapping and Spatial Analytics. Each of these packages has it's strength and weaknesses: 

`sp` is good for spatial data manipulation

`sf` is good for vector data processing and works well with tidyverse and ggplot2 

`raster` is good for raster data (imagery) processing and allows for Geodata Extraction and Manipulation

`rnaturalearth` focuses more on vector data and we've used it to download country administrative boundaries

`gisr` is a home grown package that we've put together to organize some of the most recurrent geo-related functions


```{r}

# Grand father of all Spatial Data Packages: https://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf
install.packages("sp")

# Simple Feature for R. An improve version of sp: https://r-spatial.github.io/sf/
install.packages("sf")

# Spatial Data manipulation: https://rspatial.org/raster/index.html#
install.packages("raster")

# Good resource for vector and raster data: http://www.naturalearthdata.com/
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")

# SI Packages for Mapping and Spatial Analytics: https://github.com/USAID-OHA-SI/gisr
devtools::install_github("USAID-OHA-SI/gisr")

```


## Spatial Data Acquisition


Let's download geodata from raster package

- `GADM`: vector data
- `worldclim`: raster data

```{r}

library(glamr)
library(raster)

# Setup folder
glamr::folder_setup(folder_list = list("GIS"))

# Download Admins boundaries

## National Boundaries
adm0 <- getData("GADM", country = "CIV", level = 0, download = TRUE, path = "GIS")

adm0

## Regions / province
adm1 <- getData("GADM", country = "CIV", level = 1, download = TRUE, path = "GIS")

adm1

```


Let's download geodata from rnaturaleath


```{r}

library(rnaturalearth)


# Download Admins boundaries

## National Boundaries
civ0 <- ne_countries(country = "Ivory Coast", scale = "medium", returnclass = "sf")

civ0

## Regions / province
civ1 <- ne_states(country = "Ivory Coast", returnclass = "sf")

civ1

```


Can you download geodata with `gisr` paackage?

```{r}

library(gisr)

# from rnaturalearth
get_admin0(...)
get_admin1(...)

# from raster package
get_adm_boundaries(...)

```


Let's read geodata from local file


```{r}
library(sf)

country <- read_sf("GIS/gadm36_CIV_shp/gadm36_CIV_0.shp")

country

```


## Spatial Data Visualization

There are multiple ways to visualize geodata: `plot` or `ggplot` or `gisr`

```{r}

library(tidyverse)
library(sf)
library(gisr)

# plotting sp data
plot(adm0, main = "Cote d'Ivoire")

# plotting with ggplot
regions <- st_as_sf(adm1)
regions

regions <- regions %>% 
  dplyr::select(name = NAME_1)

ggplot(data = regions) +
  geom_sf(aes(fill = name))

ggplot(data = regions) +
  geom_sf(aes(fill = name), color = "white", lwd = .2, show.legend = FALSE) +
  labs(title = "Cote d'Ivoire - Regional sub-divisions") +
  theme_void()
  
```

## Spatial Data Visualization + PEPVAR OVC Results

Now let's try to integrate the spatial dimension of PEPFAR OVC Results

Background: 

The CAC received a request for mapping support from PEDs/OVC.
The request came in a bit late and the output was supposed to be for the MOZ Deep Dive. 

[lastmile Repo](https://github.com/USAID-OHA-SI/lastmile)

1) FY20 Q2 - OVC_HIVSTAT_POS Results

```{r}
library(tidyverse)
library(sf)
library(gisr)
library(glitr)
library(patchwork)
library(here)


# Vars
country = "Mozambique"

file_psnu_txt <- "MER_Structured_Datasets_PSNU_IM_FY18-20_20200626_v2_1_Mozambique"
file_districts <- "Mozambique_PROD_5_District_DistrictLsib_2020_Feb.shp"

# Data

## Geo - Moz PSNUs Boundaries
moz_districts <- list.files(
        path = here("../../GEODATA", "PEPFAR"),
        pattern = file_districts,
        recursive = TRUE,
        full.names = TRUE
    ) %>%
    unlist() %>%
    first() %>%
    read_sf()

moz_districts %>% glimpse()

## Get admin1 geodata
moz1 <- gisr::get_admin1(country)

moz1 %>% glimpse()
    
## MER PSNU x IMs
df_psnu <- read_rds(here("../../MERDATA", paste0(file_psnu_txt, ".rds")))

df_psnu %>% glimpse()

## Filter data by indicator
moz_psnu <- df_psnu %>% 
  filter(
      fundingagency == "USAID",
      operatingunit == country,
      fiscal_year == 2020,
      indicator == 'OVC_HIVSTAT_POS'
  ) %>%
  dplyr::select(fiscal_year, snu1, psnu, psnuuid, primepartner, mech_code, mech_name, cumulative)

## Join MER data to Spatial Data
df <- moz_districts %>%
    dplyr::left_join(moz_psnu, by = c("uid" = "psnuuid")) %>%
    dplyr::filter(!is.na(cumulative))

df %>% head()

## Generate a bar chart
pos_bar <- df %>%
    dplyr::mutate(label = paste0(psnu, " (", cumulative, ")")) %>%
    ggplot2::ggplot(
        aes(x = reorder(label, cumulative),
            y = cumulative,
            fill = cumulative
        )
    ) +
    ggplot2::geom_col(show.legend = F) +
    ggplot2::scale_fill_viridis_c(option = "D", direction = -1) +
    ggplot2::scale_y_continuous(position = "right") +
    ggplot2::labs(x = "", y="") +
    ggplot2::coord_flip() +
    glitr::si_style_nolines() +
    ggplot2::theme(panel.grid.major.x = element_line(size = .2, color = grey10k))

## Generate a map 
pos_m <- gisr::terrain_map(countries = country, 
                           terr_path = "../../../GEODATA/RASTER", 
                           mask = TRUE) +
    ggplot2::geom_sf(data = df, aes(fill = cumulative), lwd = .1, color = grey10k) +
    ggplot2::geom_sf(data = moz1, fill = NA, linetype = "dotted") +
    ggplot2::geom_sf_text(data = moz1, aes(label = name), color = grey80k, size = 3) +
    ggplot2::scale_fill_viridis_c(direction = -1, na.value = grey30k) +
    ggplot2::coord_sf() +
    gisr::si_style_map() +
    ggplot2::theme(
        legend.position =  c(.9, .2),
        legend.direction = "vertical",
        legend.key.width = ggplot2::unit(.5, "cm"),
        legend.key.height = ggplot2::unit(1, "cm")
    )

## Combine bars and map + title
(pos_m + pos_bar) +
    patchwork::plot_layout(widths = c(1,1)) +
    patchwork::plot_annotation(
        title = paste0(toupper(country), " - OVC_HIVSTAT_POS Results"),
        subtitle = "Districts hightlighted in grey not part of the PSNUs",
        caption = paste0("OHA/SIEI - DATIM 2020 PSNU x IMs Data, procuded on ", Sys.Date())
    )

## Do the above with gisr function

## Combine Results + Map
spdist_ovc_hivstat_pos(country = country, 
                        fy = 2020, 
                        df_psnu = df_psnu, 
                        geo_psnu = moz_districts)

## Combine Results + Map
spdist_ovc_hivstat_pos(country = country, 
                        fy = 2020, 
                        df_psnu = df_psnu, 
                        geo_psnu = moz_districts, 
                        terr_path = "../../../GEODATA/RASTER")


```



2) FY20 Q2 - OVC TX_CURR Achievements


```{r}
library(tidyverse)
library(sf)
library(gisr)
library(glitr)
library(patchwork)
library(here)


# Vars
country = "Mozambique"

file_psnu_txt <- "MER_Structured_Datasets_PSNU_IM_FY18-20_20200626_v2_1_Mozambique"
file_districts <- "Mozambique_PROD_5_District_DistrictLsib_2020_Feb.shp"

# Use gisr to plot achievements
spdist_ovc_tx_curr(country = country,
                   fy = 2020,
                   df_psnu = df_psnu,
                   geo_psnu = moz_districts,
                   terr_path = "../../../GEODATA/RASTER")
```