---
title: "ggplot + MER"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

Today we will be diving into the some exploratory analysis with R, getting familiar with visualizing data using `ggplot2` which is part of the [tidyverse](https://www.tidyverse.org/).

To get started, you'll need to install a few different packages to work in R. Unlike proprietary software like Excel, Stata, SPSS, etc. where all the functions are built into shipped software, R is open source and allows on contributers to develop and share useful code in "packages." 

The main way to install software in R is to use `install.package()`. This function (built into R) allows you access and install software from [CRAN](https://cran.r-project.org/). 


```{r install}
install.packages("tidyverse")
install.packages("devtools")
install.packages("skimr")
install.packages("scales")
install.packages("extrafont")
install.packages("patchwork")
```

Users can also access additional packages not available on CRAN for various reasons (still in development, small scale, etc). We will also be using a package developed for use with the PEPFAR MSD, called [ICPIutilities](https://github.com/ICPI/ICPIutilities/blob/master/README.md) and is able to be installed from GitHub (via the `devtools` package).

```{r install_github}
devtools::install_github("ICPI/ICPIutilities")
```

Let's take a look at all the packages loaded with `install.packages("tidyverse")`, which actually installs a host of packages for you to use. We can see what those are through running the code below.

```{r tidyverse}
tidyverse::tidyverse_packages()
```

Great. We have the options to utitlize these packages now that they are installed,  Every time we start up a new R session, we have load the packages we want to use, using the function, `library()`.

```{r load, echo=FALSE}
library("tidyverse")
library("devtools")
library("skimr")
library("scales")
library("extrafont")
library("patchwork")
```


## Loading the data

You can use lots of different types of data in R like csv, txt, and json or even proprietary formats like xlsx and .dta. 

Today we'll be using a csv file from our training dataset that has been filtered down to a manageable size, focusing on one indicator in one operating unit.

To load the data, we will rely on a packge from the tidyverse called `readr`. One of the cool features is we can actually load data from a url and we don't need it stored locally (although, please don't store any real PEPFAR data publically).

Today's dataset is available on [Github](https://github.com/USAID-OHA-SI/coRps/blob/master/2020-03-30/FY20_MilkyWay_Testing.csv). Let's load the data locally.

```{r}
url <- "https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-03-30/FY20_MilkyWay_Testing.csv"
df_hts <- read_csv(url)
```

## Exploring the data

A best practice whenever we import a new dataset is to first inspect it. So, let's take a look.

```{r}

#high level view
  glimpse(df_hts)

#what modalities do we have?
  distinct(df_hts, modality)
  
#data distribution
  skim(df_hts, HTS_TST, HTS_TST_POS, Positivity)

```

Now that we have sense of the dataset, let start exploring the data through `ggplot`. We are model our work after the structure of R for Data Science.

## Aesthetics

```{r}
 
#dot plot
  ggplot(df_hts) +
  geom_point()

#aes in ggplot
  ggplot(df_hts, aes(HTS_TST_POS, Positivity))
  
#what is missing?
  ?geom_point

#adding a shape to compare index to non index
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, shape = is_index))
  
#adding a shape and color to index/non-index
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, shape = is_index, color = is_index))

#color by modality
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, color = modality)) 
  
#color by modality, size dot by HTS_TST
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, color = modality, size = HTS_TST))
  
#adjust transparency
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, alpha = Positivity))
    
#how do we make all the points blue?
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = "blue"))
  
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST), color = "blue")
  
#adjust dot size
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST), size = 4)
  
#adjust dot size and tranparency
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST), size = 4, alpha = .3)
  
#add mean lines to plot
  mean_hts_pos <- mean(df_hts$HTS_TST_POS)
  mean_positivity <- mean(df_hts$Positivity)
  
  ggplot(df_hts) +
  geom_vline(aes(xintercept = mean_hts_pos)) +
  geom_hline(aes(yintercept = mean_positivity)) +
  geom_point(aes(HTS_TST_POS, Positivity))
```

## Facets

```{r}

#facetting (wrap)
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality, alpha = Positivity)) +
  facet_wrap(~modality)
  
#ordered facets (wrap)
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality, alpha = Positivity)) +
  facet_wrap(~ fct_reorder(modality, HTS_TST, sum, .desc = TRUE))
  
#facet grid
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = modality, alpha = Positivity)) +
  facet_grid(modality ~ operatingunit)
  
```



## Geometric objects

```{r}
#linear fit
  ggplot(df_hts) +
  geom_smooth(aes(HTS_TST_POS, Positivity), method='lm', formula= y~x)
    
#linear fit + points (multiple layers)
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST)) +
  geom_smooth(aes(HTS_TST_POS, Positivity), method='lm', formula= y~x)
  
#linear fit by operatingunit
  ggplot(df_hts) +
  geom_point(aes(HTS_TST_POS, Positivity, size = HTS_TST, color = operatingunit)) +
  geom_smooth(aes(HTS_TST_POS, Positivity, color = operatingunit), method='lm', formula= y~x, se = FALSE)

#simplifying w/ global aes
  ggplot(df_hts, aes(HTS_TST_POS, Positivity)) +
  geom_point(aes(size = HTS_TST)) +
  geom_smooth(method='lm', formula= y~x)
```

## Transformations
```{r}
ggplot(df_hts) +
  geom_bar(aes(operatingunit))

ggplot(df_hts) +
  geom_bar(aes(operatingunit, HTS_TST_POS), stat = "identity")


ggplot(df_hts) + 
  stat_summary(
    mapping = aes(x = operatingunit, y = HTS_TST_POS),
    fun.min = min,
    fun.max = max,
    fun = median
  )


ggplot(df_hts) +
  geom_col(aes(operatingunit, HTS_TST_POS))


ggplot(df_hts) +
  geom_col(aes(operatingunit, HTS_TST_POS, fill = operatingunit))

ggplot(df_hts) +
  geom_col(aes(operatingunit, HTS_TST_POS, fill = modality), position = "dodge")

ggplot(df_hts) +
  geom_col(aes(operatingunit, HTS_TST_POS, position = "dodge", fill = operatingunit))


ggplot(df_hts) +
  geom_point(aes(Positivity, operatingunit), size = 5, alpha = .4)

ggplot(df_hts) +
  geom_point(aes(Positivity, operatingunit, color = is_index), size = 5, alpha = .4, position = "jitter")

ggplot(df_hts) +
  geom_jitter(aes(Positivity, operatingunit, color = is_index), size = 5, alpha = .4, height = .15)

```


## Coordinate systems

```{r}

library(map_data)

ggplot(df_hts) +
  geom_col(aes(modality, HTS_TST))

ggplot(df_hts) +
  geom_col(aes(modality, HTS_TST)) +
  coord_flip()

md <- map_data('county', 'marland')

ggplot(mi, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill="lightgray", colour = "white") +
  coord_quickmap()


bar <- ggplot(df_hts) + 
  geom_bar(
    mapping = aes(x = modality, fill = modality), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

  bar + coord_flip()
  bar + coord_polar()


```


   



