---
title: "ggplot + MER"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

Today we will be diving into the some exploratory analysis with R, getting familiar with visualizing data using `ggplot2` which is part of the [tidyverse](https://www.tidyverse.org/).

To get started, you'll need to install a few different packages to work in R. Unlike proprietary software like Excel, Stata, SPSS, etc. where all the functions are built into shipped software, R is open source and allows on contributers to develop and share useful code in "packages." 

The main way to install software in R is to use `install.package()`. This function (built into R) allows you access and install software from [CRAN](https://cran.r-project.org/). 


```{r install}
install.packages("tidyverse")
install.packages("devtools")
install.packages("skimr")
install.packages("scales")
install.packages("extrafont")
install.packages("patchwork")
```

Users can also access additional packages not available on CRAN for various reasons (still in development, small scale, etc). We will also be using a package developed for use with the PEPFAR MSD, called [ICPIutilities](https://github.com/ICPI/ICPIutilities/blob/master/README.md) and is able to be installed from GitHub (via the `devtools` package).

```{r install_github}
devtools::install_github("ICPI/ICPIutilities")
```

Let's take a look at all the packages loaded with `install.packages("tidyverse")`, which actually installs a host of packages for you to use. We can see what those are through running the code below.

```{r tidyverse}
tidyverse::tidyverse_packages()
```

Great. We have the options to utitlize these packages now that they are installed,  Every time we start up a new R session, we have load the packages we want to use, using the function, `library()`.

```{r load, echo=FALSE}



```


## Loading the data

You can use lots of different types of data in R like csv, txt, and json or even proprietary formats like xlsx and .dta. 

Today we'll be using a csv file from our training dataset that has been filtered down to a manageable size, focusing on one indicator in one operating unit.

To load the data, we will rely on a packge from the tidyverse called `readr`. One of the cool features is we can actually load data from a url and we don't need it stored locally (although, please don't store any real PEPFAR data publically).

Today's dataset is available on [Github](https://github.com/USAID-OHA-SI/coRps/blob/master/2020-03-30/FY20_MilkyWay_Testing.csv). Let's load the data locally.

```{r}
url <- "https://raw.githubusercontent.com/USAID-OHA-SI/coRps/master/2020-03-30/FY20_MilkyWay_Testing.csv"


```

## Exploring the data

A best practice whenever we import a new dataset is to first inspect it. So, let's take a look.

```{r}

#high level view

#what modalities do we have?

#data distribution
  skim()

```

Now that we have sense of the dataset, let start exploring the data through `ggplot`. We are model our work after the structure of R for Data Science.

## Aesthetics

```{r}
 
#dot plot
  

#aes in ggplot
  
  
#what is missing?
  

#adding a shape to compare index to non index
 
  
#adding a shape and color to index/non-index
  

#color by modality
 

#color by modality, size dot by HTS_TST
 

#adjust transparency
 
    
#how do we make all the points blue?
  

#adjust dot size
  
  
#adjust dot size and tranparency
  
  
#add mean lines to plot
 

```

## Facets

```{r}

#facetting (wrap)
  
  
#ordered facets (wrap)

  
#facet grid

  
```



## Geometric objects

```{r}
#linear fit
  



```

## Transformations
```{r}

```


 
  





